---
title: "Prelimanary Data Exloratory Data Analysis"
subtitle: "Final Project"
author: ""
date: "11/09/24"
output: html_notebook
---

**Guiding Question**
Research Question: "Is there a correlation between advanced performance metrics and award shares among players in professional basketball?"


```{r}

library(dplyr)
library(ggplot2)
library(tidyr)

```

```{r}

head(Advanced)
head(Player_Award_Shares)

```

### Merging the data
```{r}
merged_data <- merge(Advanced, Player_Award_Shares, by = c("player_id", "season"))
head(merged_data)
```

### Data Wrangling (Joins for multiple data tables)
```{r}
Advanced$player_id <- as.character(Advanced$player_id)
Advanced$season <- as.character(Advanced$season)

Player_Award_Shares$player_id <- as.character(Player_Award_Shares$player_id)
Player_Award_Shares$season <- as.character(Player_Award_Shares$season)

merged_data_join <- full_join(Advanced, Player_Award_Shares, by = c("player_id", "season"), relationship = "many-to-many")
glimpse(merged_data_join)

```

### Data Wrangling (Spread and Gather to stack/unstack variables)
```{r}
reshaped_data <- merged_data %>% 
  pivot_longer(cols = c(ws, obpm, dbpm, usg_percent), names_to = "metric", values_to = "value")
glimpse(reshaped_data)

```

### Data Wrangling (# User-defined function to categorize players by win shares)
```{r}
categorize_ws <- function(ws) {
  if (ws > 10) {
    return("High")
  } else if (ws > 5) {
    return("Medium")
  } else {
    return("Low")
  }
}
merged_data <- merged_data %>% 
  mutate(ws_category = sapply(ws, categorize_ws))

head(merged_data)

```

### Data Wrangling (Data Verbs)
```{r}
season_summary <- merged_data %>%
  group_by(season) %>%
  summarise(total_share = sum(share, na.rm = TRUE))


head(season_summary)
```

### Data Wrangling (Regular Expresison)
```{r}
mvp_data <- merged_data %>% 
  filter(grepl("MVP(\\s\\w+)?(\\s\\d{4})?", award, ignore.case = TRUE))

head(mvp_data)
```

### Data Visulazation #1
```{r}

ggplot(merged_data, aes(x = ws, y = share)) +
  geom_point() +
  geom_smooth(method = "lm", color = "red") +
  labs(title = "Win Shares vs Award Share with Smoother",
       x = "Win Shares",
       y = "Award Share")
```
### Data Visualization #2
```{r}
ggplot(merged_data, aes(x = obpm, y = share, color = ws_category)) +
  geom_point() +
  labs(title = "Offensive BPM vs Award Share by Win Share Category",
       x = "Offensive Box Plus/Minus",
       y = "Award Share",
       color = "Win Share Category")
```
### Data Visulazation #3
```{r}
ggplot(merged_data, aes(x = dbpm, fill = ws_category)) +
  geom_density(alpha = 0.4) +
  geom_point(aes(y = share / 10), color = "black", position = position_jitter(width = 0.5)) +
  labs(title = "Density of Defensive BPM by Win Share Category with Points",
       x = "Defensive Box Plus/Minus",
       y = "Density / Points")
```




